---
title: "01-Trying some ODE solvers in R"
date: 17-07-2020
---

# Trying some ODE solvers in R
```{r,warning=F,echo=FALSE}
if (!require("checkpoint", character.only = TRUE)) {
  install.packages("checkpoint", dependencies = TRUE)
  library(checkpoint)
}

if (!require("fpp", character.only = TRUE)) {
  install.packages("fpp", dependencies = TRUE)
  library(fpp)
}

dir.create(".checkpoint")
checkpoint('2020-07-17',checkpointLocation=getwd())


# Packages
library(tidyverse)
library(JuliaCall)
library(diffeqr)
library(odeintr)
library(sundialr)
library(knitr)
library(bookdown)
```
The last time I had exclusively programmed in R, it was 2014. At that point, I decided to make a shift from R to MATLAB partly because it was the tool that was used in our lab for solving ODEs and numerical analysis and secondly, R had options I was not very interested to use. Now that I have all this new found time after my pandemic defense, what better way to re-educate, and recalibrate myself and check in at what the R world is upto. 

So promptly I visited the [CRAN website]( https://cran.r-project.org/web/views/DifferentialEquations.html) and picked out the ones I found interesting to try out. Also see the [nice blog](https://www.stochasticlifestyle.com/comparison-differential-equation-solver-suites-matlab-r-julia-python-c-fortran/) comparing different solvers across different languages. As I usually work with ODEs, I checked out a standard non-linear ODE function that I wanted to integrate i.e. the Lokta-Volterra system. I also decided to not use the odeSolve package as I found the others more appealing in terms of model definition and what they interface with. 

## diffeqr package
What appealed to me is that this package interfaces with the [DifferentialEquations.jl](https://github.com/SciML/DifferentialEquations.jl) in Julia which includes solvers for DDE, ODEs, SDEs, PDEs and many more.  Moreover, in comparison to other tools, the julia package has appealing features, of which I would like to highlight event-handling, good efficiency and symbolic calculation of jacobians and autodifferentiation. Also note, that this gives us access to all the [solvers in Julia](https://docs.sciml.ai/stable/solvers/ode_solve/) Therefore, an R package that interfaces to the julia package intrgigued me. 

We begin by setting up the julia interface to R by calling `julia_setup` function in [JuliaCall](https://cran.r-project.org/web/packages/JuliaCall/index.html) package. Note that this takes some (considerable) time to run when doing it the first time.

```{r}
# The set up takes time, (a LOT of time when doing the first stime.)
julia_setup(JULIA_HOME="/Applications/Julia-1.2.app/Contents/Resources/julia/bin")
```

Then, we setup the `diffeqr` package:
```{r}
diffeqr::diffeq_setup()
```

We begin by defining our ODE system to be evaluated with Julia as 'f' and define the initial conditionas and the time span we will integrate the system over. 
```{r}
f <- JuliaCall::julia_eval("
function f(du,u,p,t)
  du[1] = 10.0*(u[2]-u[1])
  du[2] = u[1]*(28.0-u[3]) - u[2]
  du[3] = u[1]*u[2] - (8/3)*u[3]
end")

u0 = c(1.0,0.0,0.0)
tspan <- list(0.0,100.0)

```

We then solve the ODE. Note that the first time we called the model it also compiles, therefore I store only the second time taken in the second time we solve it. 
```{r}

for(i in 1:2){
t_diffeqr<-system.time(
  {sol = diffeqr::ode.solve('f',u0,tspan,reltol=1e-4,abstol=1e-4)}
  )
}
colnames(sol$u)<-c('u1','u2','u3')
sol_df=sol$u%>%
      as_tibble()%>%
      mutate(time=sol$t)

sol_diffeqr_melted=sol_df%>%
  gather(key=u,value='value',-time)%>%
  mutate(Method='diffeqr')
```

## odeintr package
The odeintr package uses the integration engine from [Boost](https://headmyshoulder.github.io/odeint-v2/) and therefore compiles the model in C++. We specify the model in C, compile and integrate it. The nice part it this package allows you to use all the steppers in `Boost`. I liked the intuitive/ clean way of defining the models and running it. 

```{r}
f_c <- '
  dxdt[0] = 10.0*(x[1]-x[0]);
  dxdt[1] = x[0]*(28.0-x[2]) - x[1];
  dxdt[2] = x[0]*x[1] - (8/3)*x[2];
'

# compile model, and pass integration options
compile_sys("lorenz", f_c,atol = 1e-04, rtol = 1e-04)

# Run model
t_odeintr=system.time({sol_odeintr = lorenz(u0, 100, 0.001)})

# Results
colnames(sol_odeintr)<-c('time','u1','u2','u3')
sol_odeintr_melted<-sol_odeintr%>%
  as_tibble()%>%
  gather(key=u,value='value',-time)%>%
  mutate(Method='odeintr')

```

## sundialr package
Finally, the [subdialr](https://cran.r-project.org/web/packages/sundialr/index.html) package that interfaces with CVODE/CVODES (something I am more familiar with). I define a model in R and run it: 

```{r}
f_sundials<- function(t,u,p){
  du <- vector(mode = "numeric", length = length(u))
  
  du[1] = 10.0*(u[2]-u[1])
  du[2] = u[1]*(28.0-u[3]) - u[2]
  du[3] = u[1]*u[2] - (8/3)*u[3]
  
  du
}

# run
t_cvode=system.time({
sol_cvode<- cvode(seq(0,100,0.01), u0+1e-10, f_sundials,0,reltolerance = 1e-4,abstolerance = 1e-4)
})

# 
colnames(sol_cvode)<-c('time','u1','u2','u3')
sol_cvode_melted<-sol_cvode%>%
  as_tibble()%>%
  gather(key=u,value='value',-time)%>%
  mutate(Method='sundialr')

```

and also define one with RCpp: 
```{r}
# With RCpp
Rcpp::sourceCpp(code = '
#include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
NumericVector f_sundials_Rcpp (double t, NumericVector y, NumericVector p){
  // Initialize ydot filled with zeros
  NumericVector dydt(y.length());
  // 
   dydt[0] = 10.0*(y[1]-y[0]);
  dydt[1] = y[0]*(28.0-y[2]) - y[1];
  dydt[2] = y[0]*y[1] - (8/3)*y[2];
  return dydt;
}')

# run
t_cvode_cpp=system.time({
  sol_cvode_cpp<- cvode(seq(0,100,0.01), u0+1e-10, f_sundials_Rcpp, NaN,reltolerance = 1e-4,abstolerance = 1e-4)
})

# 
colnames(sol_cvode_cpp)<-c('time','u1','u2','u3')
sol_cvode_cpp_melted<-sol_cvode_cpp%>%
  as_tibble()%>%
  gather(key=u,value='value',-time)%>%
  mutate(Method='sundialr_Rcpp')

```

## Some remarks
Taking a quick look at the CPU times shows that the `diffeqr` and `odeintr` package are faster than the `sundialr` one. 

```{r,warning=FALSE}
all_sols=bind_rows(sol_cvode_melted,sol_cvode_cpp_melted,
                   sol_odeintr_melted,sol_diffeqr_melted)
ggplot(all_sols,aes(x=time,y=value))+
  geom_line(aes(col=u))+
  scale_color_brewer(palette='Set1')+
  theme_minimal()+
  facet_wrap(~Method)

# Time 
time_df=bind_rows(t_cvode_cpp,t_cvode,t_diffeqr,t_odeintr);
time_df<-time_df%>%mutate(package=c('sundialr_Cpp',
                                    'sundialr',
                                    'diffeqr',
                                    'odeintr'))

kable(time_df)
```


Finally, 'deSolve' also allows us to interface with FORTRAN and C. Since its uses LSODA for solving ODEs. However, I decided against trying this, given the superiority of the methods available in `DifferentialEquations.jl`. Since a lot of packages are based on 'deSolve' I think it deserves a separate and a more serious comparison with the other packages in terms of efficiency, choice of methods available and usability. 


